Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-08 paper/ms.tex
update validation plot annotations
first draft of Fourier text, tightened up a bit of language, and added some citations

@@

%\title{Pre-Main-Sequence M-Dwarfs With Complex Quasiperiodic Variability From Four Years of TESS}
%\title{Complex Quasiperiodic Variables From Four Years of TESS}
\title{Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of TESS}

@@

\shortauthors{Bouma, Jayararaman, et al.}

@@

\author[0000-0002-7778-3117]{Rahul~Jayararaman}
\affiliation{\mitkavli}

@@

Data from K2 and TESS have yielded a new class of variable star whose
root cause is only beginning to become clear: complex quasiperiodic
variables (CQVs).  These objects are identified phenomenologically
using their optical light curves, which show nearly periodic troughs
that can be either sharp or broad, often superposed on smooth
spot-like modulation
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight local minima (dips) per cycle.  Most are
pre-main-sequence M dwarfs without near-infrared excesses, ages of
$\approx$5-150 million years (Myr), and rotation periods of at most
two days; they are observed to be $\approx$1-3\% of M dwarfs younger
than 100 Myr \citep{2016AJ....152..114R,2022AJ....163..144G}.  The
dips can be chromatic, with a reddening law plausibly consistent with
dust
\citep{2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
And finally, while the dip shapes can ``jump'' between different
depths and durations over less than one cycle, they more often evolve
gradually, over tens to hundreds of cycles
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

@@

CQVs remain mysterious because they have been both hard to discover
and hard to characterize.   Discoverability is tied to rarity: CQVs
comprise $\approx$1\% of the youngest $\approx$1\% of M dwarfs
\citep{2018AJ....155..196R}.  Out of the millions of stars monitored
by K2 and TESS, about 50 CQVs have been reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
The known CQVs are correspondingly faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  This renders time-series spectroscopy
at high resolution out of reach for current facilities, despite its
potential utility in ruling between the models.

@@

A point on nomenclature.  CQVs have been called ``transient flux
dips'', ``persistent flux dips'', ``scallop shells'', ``batwings'',
\citep{2017AJ....153..152S} ``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P}
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and large-amplitude variability
linked to obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  With that said, a
few similarities between CQVs and dippers do exist (see
Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  At the
risk of introducing yet another standard, we hope to introduce a
nomenclature that reflects how, when observed over timescales of more
than tens of cycles, CQVs are almost but not exactly periodic.  They
are quasiperiodic.  While the three-type classification scheme
proposed by \citet{2017AJ....153..152S} may indeed provide some
helpful visual distinctions amongst the CQVs, it seems likely that
they are all explained by a single underlying phenomenon, and so we
opt to refer to them by a single empirically descriptive name.

@@

We analyzed the ``short'' 120-second cadence data acquired by TESS
between July~2018 and Sep~2022 (Sectors 1-55).  Specifically, we used
the 120-second cadence light curve products produced by the 
Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products also
include full frame images with cadences of 200, 600, and 1800 seconds,
we limited our scope in this work for the sake of simplicity in data
handling.  In exchange, we lose in both completeness and homogeneity
of the selection function.  While TESS cumulatively observed
$\approx$90\% of the sky for at least one lunar month between
July~2018 and Sep~2022, the 120-second cadence data were acquired for
only a pre-selected set of stars over Sectors 1-26, and then a
guest-investigator driven set of stars over Sectors 27-55
\citep{2021PASP..133i5002F}.

@@

\subsubsection{Fourier analysis}
\label{subsec:fourier}
For the Fourier analysis, we followed \citet{2019ApJ...876..127Z}.

{\bf TODO for Rahul or Saul: explain the approach, in a few
	paragraphs.  Was the SAP\_FLUX or PDCSAP used? etc. }

@@

Broadly speaking, the most common false positives for both the Fourier
and dip-counting techniques were eclipsing binaries, spot-induced
variability from rapid rotators, and variability from neighboring,
off-target stars.  Typical false positive rates from our dip-counting
pipeline were 5:1, with \nuniqdipflagged\ unique stars flagged, and
about 20\% being labeled either ``good'' or ``possible'' CQVs; for
the Fourier pipeline, {\bf the rate was X:Y, with ZZZZ unique stars
flagged, and NN\% being labeled ``good'' or ``possible'' CQVs}.

@@

An argument against the connection between CQVs and the
$\sigma$~Ori~E analogs is that the light curve of $\sigma$~Ori~E is
much simpler than those in Figure~\ref{fig:cqvs}, with only two broad
local minima, and one ``hump'' \citep{2005ApJ...630L..81T}.  Within
the model proposed by Townsend et al., this is explained by the simplicity of the star's
magnetic field, which to first-order is a
dipole, similar to most other magnetic B stars
\citep{2009ARA&A..47..333D}.  This approximation matches the qualitative
behavior of both the broadband light curves, and the
line-profile variations seen in hydrogen, helium, and carbon
\citep{2012MNRAS.419..959O}.

Two exceptions to the rule of ``simple field topologies for magnetic
B stars'' are HD~37776 and HD~64740, which are B stars that were
previously known from spectropolarimetry in the 2000s-2010s to have
field geometries dominated by high order multipoles
\citep{2011ApJ...726...24K}.  In a review of the literature, we
learned that recent TESS light curves of these two B stars appear
surprisingly similar to the CQV light curves
\citep{2020pase.conf...46M}.  Figure~\ref{fig:bstar} shows the phased
TESS light curves for the two outlying magnetic B stars, with by-eye
``best matching'' CQVs shown for comparison.  The number of dips per
cycle, the shapes of the dips, and the dip depths relative to the
sinusoidal envelope, all seem to be quite similar.  If this connection
is taken seriously, the implication would be that the
highly-structured M dwarf light curves imply multipolar magnetic
fields -- since it is the complex fields which are the defining
characteristic of these two B stars, relative to other known magnetic
B stars like $\sigma$~Ori~E.  
This is consistent with the non-axisymmetric topologies that have been
reported in a small number of M dwarfs for which Zeeman Doppler
Imaging is technically feasible 
\citep[see the review by][]{2021A&ARv..29....1K}.

@@

\software{
  %\texttt{arviz} \citep{arviz_2019},
  %\texttt{altaipony} \citep{ilin_flares_2021},
  \texttt{astrobase} \citep{2021zndo...1011188B},
  %\texttt{astroplan} \citep{astroplan2018},
	%\texttt{AstroImageJ} \citep{collins_astroimagej_2017},
  %\texttt{astropy} \citep{astropy_2018},
  %\texttt{astroquery} \citep{astroquery_2018},
  %\texttt{BATMAN} \citep{kreidberg_batman_2015},
  %\texttt{ceres} \citep{brahm_2017_ceres},
  %\texttt{cdips-pipeline} \citep{bhatti_cdips-pipeline_2019},
  %\texttt{corner} \citep{corner_2016},
  %\texttt{emcee} \citep{foreman-mackey_emcee_2013},
  %\texttt{exoplanet} \citep{exoplanet:exoplanet}, and its
  %dependencies \citep{exoplanet:agol20, exoplanet:kipping13, exoplanet:luger18,
  % 	exoplanet:theano},
	%\texttt{gala} \citep{gala,PriceWhelan_2017_gala_zenodo},
	%\texttt{IDL Astronomy User's Library} \citep{landsman_1995},
  %\texttt{IPython} \citep{perez_2007},
	%\texttt{isochrones} \citep{morton_2015_isochrones},
	\texttt{lightkurve} \citep{2018ascl.soft12013L},
  %\texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  %\texttt{MESA} \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  %\texttt{numpy} \citep{walt_numpy_2011}, 
  %\texttt{pandas} \citep{mckinney-proc-scipy-2010},
  %\texttt{pyGAM} \citep{serven_pygam_2018_1476122},
  %\texttt{PyMC3} \citep{salvatier_2016_PyMC3},
  %\texttt{radvel} \citep{fulton_radvel_2018},
  %\texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{2020NatMe..17..261V},
  \texttt{TESS-point}  \citep{2020ascl.soft03001B},
  %\texttt{tesscut} \citep{brasseur_astrocut_2019},
  %\texttt{unpopular} \citep{hattorio_2021_cpm},
  %\texttt{VESPA} \citep{morton_efficient_2012,vespa_2015},
  %\texttt{webplotdigitzer} \citep{rohatgi_2019},
  %\texttt{wotan} \citep{hippke_wotan_2019}.
}
\

>>>>

\newcommand{\red}{\color{red}}

@@

%\title{Pre-Main-Sequence M-Dwarfs With Complex Quasiperiodic Variability From Four Years of TESS}
%\title{Complex Quasiperiodic Variables From Four Years of TESS}
\title{Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS}

@@

\shortauthors{Bouma, Jayaraman, et al.}

@@

\author[0000-0002-7778-3117]{Rahul~Jayaraman}
\affiliation{\mitkavli}

@@

Data from K2 and TESS have yielded a new class of variable star---
complex quasiperiodic variables---for which the underlying cause of
variability has only begun to be explored recently. These objects are 
identified phenomenologically
using their optical light curves, which show nearly periodic troughs
that are either sharp or broad; these are often superposed on smooth
spot-like modulations
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight local minima (``dips'') per cycle.  Most are
pre-main-sequence M dwarfs without near-infrared excesses, ages of
$\approx$5-150 million years (Myr), and rotation periods of at most
two days; they are observed to be $\approx$1-3\% of M dwarfs younger
than 100 Myr \citep{2016AJ....152..114R,2022AJ....163..144G}.  The
dips can be chromatic, with a reddening law plausibly consistent with
dust
\citep{2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
And finally, while the dip shapes can ``jump'' between different
depths and durations over less than one cycle, they more often evolve
gradually, over tens to hundreds of cycles
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

@@

CQVs remain mysterious because they have been both hard to discover
and hard to characterize.   Discoverability is tied to rarity: CQVs
comprise $\approx$1\% of the youngest $\approx$1\% of M dwarfs
\citep{2018AJ....155..196R}.  Out of the millions of stars monitored
by K2 and TESS, about 50 CQVs have been reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
The known CQVs are correspondingly faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  This renders time-series spectroscopy
at high resolution out of reach for current facilities, despite its
potential utility in ruling out some of these models.

@@

We note that the ``proper'' nomenclature for these stars in the literature
has generally been unclear. CQVs have been called ``transient flux
dips'', ``persistent flux dips'', ``scallop shells'', ``batwings'',
\citep{2017AJ....153..152S} ``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P}
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and large-amplitude variability
linked to obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  With that said, a
few similarities between CQVs and dippers do exist (see
Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  At the
risk of introducing yet another standard, we hope to introduce a
nomenclature that reflects how, when observed over timescales of more
than tens of cycles, CQVs are almost but not exactly periodic.  They
are quasiperiodic.  While the three-type classification scheme
proposed by \citet{2017AJ....153..152S} may indeed provide some
helpful visual distinctions amongst the CQVs, it seems likely that
they are all explained by a single underlying phenomenon, and so we
opt to refer to them by a single empirically descriptive name.

@@

We analyzed the ``short'' 120-second cadence data acquired by TESS
between July~2018 and Sep~2022 (Sectors 1-55).  Specifically, we used
the 120-second cadence light curve products produced by the 
Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products from these 
sectors also include full frame images with cadences of 600 and 1800 seconds,
we limited our scope to 120-s cadence in this work for the sake of simplicity in data
handling.  In exchange, we lose in both completeness and homogeneity
of the selection function.  While TESS cumulatively observed
$\approx$90\% of the sky for at least one lunar month between
July~2018 and Sep~2022, the 120-second cadence data were acquired for
only a pre-selected set of stars over Sectors 1-26, and then a
guest-investigator driven set of stars over Sectors 27-55
\citep{2021PASP..133i5002F}.

@@

\subsubsection{Fourier analysis}
\label{subsec:fourier}
For the Fourier analysis, we followed the techniques presented in 
\citet{2019ApJ...876..127Z} and Section 1.3 of \citet{2023MNRAS.524.4220P}.
All 120-s targets were searched for strongly periodic phenomena using our
Fourier pipeline. We used the {\tt PDCSAP\_FLUX} column from the data  
and normalized and  re-binned the light curve into equal width bins, due to the
uneven spacing in the TESS data, as well as the data gap in the middle of
each sector due to the lack of data collection during downlink. Data was
padded to ensure that the light curve had a length that was a power of 2, 
as described in \citeauthor{2019ApJ...876..127Z}. We
then took the Fourier transform of the corrected, padded light curve using 
{\tt numpy.fft.fft} and searched for peaks with a significance of over 
12-$\sigma$ within each set of 500 bins.

If a peak of such significance was found, we generated a ``summary sheet'' 
with information about the target, its full and folded light curves, Fourier 
transform, potential contaminating stars, and information about these
 contaminating stars. Throughout the 35 sectors for which we performed
 this analysis, we found that ``summary'' sheets were generated for 
 approximately 10\% of the 20\,000
120-s targets in a given sector. Each summary sheet was manually reviewed.
During this manual review, each light curve was sorted into a category based on its
morphology (e.g., eclipsing binary, CQV, or cataclysmic variable). For
the CQVs, a further cut was made based on the target's temperature---any
stars exhibiting phenomena that looked similar to CQVs, but had a $T_{\rm eff}$ 
over 4\,000\,K were discarded from the target list (e.g., RS CVn stars). 
{\red (RJ: can we add an example of these sheets to this paper? i know it's not 
the ``main'' vetting, though, so happy to not do this as well)}.

CQVs that were observed in multiple sectors had their light curves
stitched together and analyzed for a more accurate period determination.

@@

Broadly speaking, the most common false positives for both the Fourier
and dip-counting techniques were eclipsing binaries, spot-induced
variability from rapid rotators, and variability from neighboring,
off-target stars.  Typical false positive rates from our dip-counting
pipeline were 5:1, with \nuniqdipflagged\ unique stars flagged, and
about 20\% being labeled either ``good'' or ``possible'' CQVs.

The Fourier pipeline, on the other hand, was less CQV-focused than
the dip-counting technique, as its results were used for other sorts of
analyses; as a result, we cannot calculate a similar ``false-positive''
probability. Between the targets identified as CQVs by both pipelines,

 {\red

\begin{itemize}
\item XX targets were found to overlap
\item YY targets were found via Fourier but not via dip counting
\end{itemize}

}
The difference in discovery rate between these techniques can be attributed to
low-amplitude peaks in the periodogram that were missed, some of which likely
arose from fainter CQVs.

@@

An argument against the connection between CQVs and the
$\sigma$~Ori~E analogs is that the light curve of $\sigma$~Ori~E is
much simpler than those in Figure~\ref{fig:cqvs}, with only two broad
local minima, and one ``hump'' \citep{2005ApJ...630L..81T, 2022ApJ...924L..10J}.  
Within the model proposed by Townsend et al., this is explained by the simplicity of 
the star's magnetic field, which to first-order is a
dipole, similar to most other magnetic B stars
\citep{2009ARA&A..47..333D}.  This approximation matches the qualitative
behavior of both the broadband light curves, and the
line-profile variations seen in hydrogen, helium, and carbon
\citep{2012MNRAS.419..959O}.

Two exceptions to the rule of ``simple field topologies for magnetic
B stars'' are HD~37776 and HD~64740, which are B stars that were
previously known from spectropolarimetry in the 2000s-2010s to have
field geometries dominated by high order multipoles
\citep{2011ApJ...726...24K}.  In a review of the literature, we
learned that recent TESS light curves of these two B stars appear
surprisingly similar to the CQV light curves
\citep{2020pase.conf...46M}.  Figure~\ref{fig:bstar} shows the phased
TESS light curves for the two outlying magnetic B stars, with by-eye
``best matching'' CQVs shown for comparison.  The number of dips per
cycle, the shapes of the dips, and the dip depths relative to the
sinusoidal envelope, all seem to be quite similar.  The
implication of such a surprising connection would be that the
highly-structured M dwarf light curves imply multipolar magnetic
fields -- since it is the complex fields which are the defining
characteristic of these two B stars, relative to other known magnetic
B stars like $\sigma$~Ori~E.  
This is consistent with the non-axisymmetric topologies that have been
reported in a small number of M dwarfs for which Zeeman Doppler
Imaging is technically feasible 
\citep[see the review by][]{2021A&ARv..29....1K}.

@@

\software{
  %\texttt{arviz} \citep{arviz_2019},
  %\texttt{altaipony} \citep{ilin_flares_2021},
  \texttt{astrobase} \citep{2021zndo...1011188B},
  %\texttt{astroplan} \citep{astroplan2018},
	%\texttt{AstroImageJ} \citep{collins_astroimagej_2017},
  \texttt{astropy} \citep{astropy_2013,astropy_2018,astropy_2022},
  %\texttt{astroquery} \citep{astroquery_2018},
  %\texttt{BATMAN} \citep{kreidberg_batman_2015},
  %\texttt{ceres} \citep{brahm_2017_ceres},
  %\texttt{cdips-pipeline} \citep{bhatti_cdips-pipeline_2019},
  %\texttt{corner} \citep{corner_2016},
  %\texttt{emcee} \citep{foreman-mackey_emcee_2013},
  %\texttt{exoplanet} \citep{exoplanet:exoplanet}, and its
  %dependencies \citep{exoplanet:agol20, exoplanet:kipping13, exoplanet:luger18,
  % 	exoplanet:theano},
	%\texttt{gala} \citep{gala,PriceWhelan_2017_gala_zenodo},
	%\texttt{IDL Astronomy User's Library} \citep{landsman_1995},
  %\texttt{IPython} \citep{perez_2007},
	%\texttt{isochrones} \citep{morton_2015_isochrones},
	\texttt{lightkurve} \citep{2018ascl.soft12013L},
  %\texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  %\texttt{MESA} \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  \texttt{numpy} \citep{2020Natur.585..357H}, 
  %\texttt{pandas} \citep{mckinney-proc-scipy-2010},
  %\texttt{pyGAM} \citep{serven_pygam_2018_1476122},
  %\texttt{PyMC3} \citep{salvatier_2016_PyMC3},
  %\texttt{radvel} \citep{fulton_radvel_2018},
  %\texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{2020NatMe..17..261V},
  \texttt{TESS-point}  \citep{2020ascl.soft03001B},
  %\texttt{tesscut} \citep{brasseur_astrocut_2019},
  %\texttt{unpopular} \citep{hattorio_2021_cpm},
  %\texttt{VESPA} \citep{morton_efficient_2012,vespa_2015},
  %\texttt{webplotdigitzer} \citep{rohatgi_2019},
  %\texttt{wotan} \citep{hippke_wotan_2019}.
}
\
